---
title: "Water inequalities across CPLP member states"
author: Sílvia Paiva
date: today
format: 
  html:
    theme: cerulean
editor: visual
---

### 1. Introduction

Although water covers most of the Earth’s surface, less than 1% is readily accessible freshwater for human use. Growing populations, environmental pressures, and uneven infrastructure have intensified global challenges in securing safe and reliable drinking water. In response, the United Nations established Sustainable Development Goal 6 (SDG 6), which aims to ensure universal access to safe water and sanitation by 2030.

Despite progress, large inequalities persist, particularly in low- and middle-income regions, where inadequate water services continue to affect health, well-being, and economic development. To better understand these disparities, this project analyses the evolution of access to drinking water in Portuguese-speaking countries (CPLP) over the past 25 years, comparing national trajectories and inequalities across residence types.

#### Data Description

This project uses drinking water access estimates from the WHO/UNICEF Joint Monitoring Programme (JMP). The raw dataset includes information for all countries from 2000 to 2024 and contains variables such as country identifiers, SDG region, income group, population, and service-level indicators (e.g., at least basic, limited, unimproved, safely managed). For this project, the data were cleaned, transformed into a long format, and restricted to the CPLP countries to enable consistent comparison across years and service levels.

#### Project goals

My main goal is to evaluate how access to drinking water has evolved in CPLP countries over the past 25 years and to quantify inequalities between rural and urban populations. A secondary goal is to compare CPLP performance with broader SDG regional trends to contextualize progress and identify areas where improvements are most needed.

### 2. Data Analysis

#### Load Packages

```{r}
#| echo: false
library(tidyverse)
library(ggthemes)
library(here)
```

#### Import Data

```{r}
#| echo: false
water <- read_csv(here::here("data/processed/water_data_long.csv"))

glimpse(water)
```

#### Data Cleaning

### 3. Results

```{r}
water_access <- water |>
  filter(varname_short == "wat_bas") |> 
  mutate(percent = as.numeric(percent),
         percent = if_else(percent < 0 | percent > 100, NA_real_, percent))

water_access |>
  filter(residence == "Total") |>
  ggplot(aes(x = year, y = percent, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Access to At Least Basic Drinking Water in CPLP Countries (Total Population)",
    x = "Year",
    y = "Percentage of Population",
    color = "Country"
  )
```

## 

### 3. Discussion & Conclusions

### 4. References
