---
title: "Water inequalities across CPLP member states"
format: html
editor: visual
---

### **1. Project Description**

Global access to safe water is central to SDG 6.

CPLP (Community of Portuguese-Speaking Countries) is a unique geopolitical group spanning Africa, South America, Asia, and Europe, with diverse socioeconomic realities.

#### **1.3 Research Questions**

i\. How do CPLP countries evolved in access to drinking water over the past 25 years?

ii\. How do rural-urban inequalities vary across the CPLP?

iii\. How do improvement in water services correlate with improvements in sanitation and hygiene?

iv\. How do CPLP countries compare to their broader SDG regional averages?

v\. Where are the largest gaps, fastest improvements, and critical vulnerabilities?

**Inequality and vulnerability profiles**

What indicators drive the largest disparities?

Which CPLP countries have the worst rural disadvantages? Which countries progressed fastest?

### Load Packages

```{r}
library(tidyverse)
library(ggthemes)
library(here)
```

## Import Data

```{r}
#| echo: false
water <- read_csv(here::here("data/processed/water_data_long.csv"))

# Inspect
glimpse(water)
```

## Data Cleaning

#### Cleaning Water Dataset

```{r}
water_access <- water |>
  filter(varname_short == "wat_bas") |> 
  mutate(percent = as.numeric(percent),
         percent = if_else(percent < 0 | percent > 100, NA_real_, percent))

water_access |>
  filter(residence == "Total") |>
  ggplot(aes(x = year, y = percent, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Access to At Least Basic Drinking Water in CPLP Countries (Total Population)",
    x = "Year",
    y = "Percentage of Population",
    color = "Country"
  )
```

## 
